Practical Optional Types for Clojure

### 2015-07-03 formal-model.tex
Add section 2 intro. Move exceptional flow to section 3 unedited.

>>>>

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants.

\begin{exmp}
\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
\label{example:doexception}
\end{exmp}

The fully expanded increment function in~\egref{example:doexception}
guards its final call with a number check, preventing
a possible null-pointer exception.
Without this check, the type system would reject the program.

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{See \url{https://github.com/typedclojure/examples}
  for full examples.}
We model this formally %(section~\ref{sec:doformal}) 
and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).

### 2015-07-03 overview.tex
More editing.
Add section 2 intro. Move exceptional flow to section 3 unedited.

\subsection{Clojure Syntax}

We describe new syntax as they appear in each example, but
we also include the essential basics of Clojure syntax.

\clj{nil} is exactly Java's \java{null}.
Parentheses indicate \emph{applications}, brackets
delimit
\emph{vectors}, braces
delimit
\emph{hash-maps}
and double quotes delimit \emph{Java strings}.
\emph{Symbols} begin with an alphabetic character,
and a colon prefixed symbol like \clj{:a} is a \emph{keyword}.

\emph{Commas} are always \emph{whitespace}.

\subsection{Typed Racket and occurrence typing}

\citet{TF10}
presented Typed Racket with occurrence typing,
a technique for deriving type information from conditional control flow.
They introduced the concept of occurrence typing 
with the following example.

\inputminted[firstline=3]{racket}{code/tr/example1.rkt}

This function takes a value that is either \emph{\#f} % mintinline really hates #
or a number, represented by an \emph{untagged} union type.
The `then' branch has an implicit invariant
that \rkt{x} is a number, which is automatically inferred with occurrence typing
and type checked without further annotations.

We chose to build on the ideas and implementation
of Typed Racket to implement a type system targeting Clojure for several reasons.
Initially, the similarities between Racket and Clojure drew us to
investigate the effectiveness of repurposing occurrence typing
for a Clojure type system---both languages share a Lisp heritage,
similar standard functions 
(for instance \clj{map}
in both languages is variable-arity)
and idioms.
While Typed Racket is gradually typed and has sophisticated
dynamic semantics for cross-language interaction, we 
chose to first implement
the static semantics
with the hope to extend Typed Clojure to be gradually typed at a future date.
Finally,
Typed Racket's combination of bidirectional checking
and occurrence typing presents a successful model for 
type checking dynamically typed programs without compromising
soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
which cannot prove strong properties about programs
and soft typing~\cite{CF91}
which has proved too complicated in practice.

Here is the above program in Typed Clojure.
\begin{exmp}
\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
\label{example:conditionalflow}
\end{exmp}

The \clj{fn} macro (provided by core.typed) supports optional annotations by 
adding
\clj{:-} and a type after a parameter
position
or binding vector 
to annotate parameter types
and return types respectively.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
any number of types.

Typed Clojure can already check all of the examples in~\citet{TF10}---the 
rest of this section describes the extensions necessary
to check Clojure code.

\subsection{Exceptional control flow}

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants.

\begin{exmp}
\inputminted[firstline=13,lastline=15]{clojure}{code/demo/src/demo/do.clj}
\label{example:doexception}
\end{exmp}

The fully expanded increment function in~\egref{example:doexception}
guards its final call with a number check, preventing
a possible null-pointer exception.
Without this check, the type system would reject the program.

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{See \url{https://github.com/typedclojure/examples}
  for full examples.}
We model this formally (section~\ref{sec:doformal}) and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).

@@

The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
The function \clj{an-exp} is verified to return an \clj{Expr}.

The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
Here \clj{defalias} defines \clj{Expr}
that describes the structure of a recursively-defined AST as a union of HMaps.
A quoted keyword in a type, such as \clj{':if}, is a singleton type that contains just the keyword.
A type that is a quoted map like \clj{'{:op ':if}} is a
HMap type with a fixed number of keyword entries of the specified types
known to be \emph{present},
zero entries known to absolutely be \emph{absent},
and an infinite number of \emph{unknown} entries entries.
Since only keyword keys are allowed, they do not require quoting.

>>>>

%\paragraph{Clojure Syntax}
%
%We describe new syntax as they appear in each example, but
%begin with include the essential basics of Clojure syntax.
%
%\clj{nil} is exactly Java's \java{null}.
%Parentheses indicate \emph{applications}, brackets
%delimit
%\emph{vectors}, braces
%delimit
%\emph{hash-maps}
%and double quotes delimit \emph{Java strings}.
%\emph{Symbols} begin with an alphabetic character,
%and a colon prefixed symbol like \clj{:a} is a \emph{keyword}.
%
%\emph{Commas} are always \emph{whitespace}.

\subsection{Typed Clojure Introduction}

Here is a simple program in Typed Clojure.

We define a global variable \clj{greet} as a one-argument function
taking and returning a string.

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n]
  (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
(greet nil)    ; Type Error:
               ;   Expected Str, given nil
\end{minted}

The function may take strings, but providing \clj{nil} (Clojure's name for Java's \emph{null})
is a static type error --- \clj{nil} is not a string in Typed Clojure.

\paragraph{Unions} We can make the annotation more permissive with \emph{ad-hoc unions}
to allow \clj{nil}.

\begin{minted}{clj}
(ann  greet-nil [(U nil Str) -> Str])
(defn greet-nil [n]
  (str "Hello" (when n (str ", " n)) "!"))
(greet-nil "Donald") ; :- Str
;=> "Hello, Donald!"
(greet-nil nil)    ; :- Str
;=> "Hello!"
\end{minted}

All Clojure values are true, except \clj{nil} and \clj{false}, so the
comma is only added when the argument is non-nil.

\paragraph{Flow analysis} Typed Clojure uses occurrence typing to
model type-based control flow.

\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, "
       (when i
         (apply str (repeat i (str "hello, "))))
       n "!"))
(greetings "Donald" 2)  ; :- Str
;=> "Hello, hello, hello, Donald!"
(greetings "Grace" nil) ; :- Str
;=> "Hello, Grace!"
\end{minted}

Providing \clj{repeat} with a nilable first argument is disallowed,
but the \clj{when} expression ensures it can only be passed an integer.

%\subsection{Type System Basics}
%
%\citet{TF10}
%presented Typed Racket with occurrence typing,
%a technique for deriving type information from conditional control flow.
%They introduced the concept of occurrence typing 
%with the following example.
%
%\inputminted[firstline=3]{racket}{code/tr/example1.rkt}
%
%This function takes a value that is either \emph{\#f} % mintinline really hates #
%or a number, represented by an \emph{untagged} union type.
%The `then' branch has an implicit invariant
%that \rkt{x} is a number, which is automatically inferred with occurrence typing
%and type checked without further annotations.
%
%We chose to build on the ideas and implementation
%of Typed Racket to implement a type system targeting Clojure for several reasons.
%Initially, the similarities between Racket and Clojure drew us to
%investigate the effectiveness of repurposing occurrence typing
%for a Clojure type system---both languages share a Lisp heritage,
%similar standard functions 
%(for instance \clj{map}
%in both languages is variable-arity)
%and idioms.
%While Typed Racket is gradually typed and has sophisticated
%dynamic semantics for cross-language interaction, we 
%chose to first implement
%the static semantics
%with the hope to extend Typed Clojure to be gradually typed at a future date.
%Finally,
%Typed Racket's combination of bidirectional checking
%and occurrence typing presents a successful model for 
%type checking dynamically typed programs without compromising
%soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
%which cannot prove strong properties about programs
%and soft typing~\cite{CF91}
%which has proved too complicated in practice.
%
%Here is the above program in Typed Clojure.
%\begin{exmp}
%\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
%\label{example:conditionalflow}
%\end{exmp}
%
%The \clj{fn} macro (provided by core.typed) supports optional annotations by 
%adding
%\clj{:-} and a type after a parameter
%position
%or binding vector 
%to annotate parameter types
%and return types respectively.
%\clj{number?} is
%a Java \java{instanceof} test of \clj{java.lang.Number}.
%As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
%any number of types.
%
%Typed Clojure can already check all of the examples in~\citet{TF10}---the 
%rest of this section describes the extensions necessary
%to check Clojure code.

@@

%The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
%The function \clj{an-exp} is verified to return an \clj{Expr}.
%
Here \clj{defalias} defines \clj{Expr}, a type abbreviation
that describes the structure of a recursively-defined AST as a union of HMaps.
A quoted keyword in a type --- \clj{':if} --- is a singleton type that contains just the keyword.
A type that is a quoted map like \clj{'{:op ':if}} is a
HMap type with a fixed number of keyword entries of the specified types
known to be \emph{present},
zero entries known to absolutely be \emph{absent},
and an infinite number of \emph{unknown} entries entries.
Since only keyword keys are allowed, they do not require quoting.
